Some examples in extended logic counting classes

### 2016-12-31 LICS/appendix-syntactic.tex
30 dic
31 dic merge

>>>>

We give this proof in various parts.

We show here that $\QE{0} \not\subseteq \E{1}$.. By contradiction, suppose that there is a $\QE{0}$ formula $\alpha$ over some signature $\R$ such that defines the following function. For every finite $\R$-structure with $n$ elements, and where every predicate in $\R$ is empty, $\alpha(\enc(\A)) = n - 1$. We use the following claim.
Let $\alpha = \sa{\bar{x}}\varphi(\bar{x})$	where $\varphi$ is quantifier free. Then the function defined by $\alpha$ is either null, greater or equal to $n$, or is in $\Omega(n^2)$.
Suppose that the function defined by $\alpha$ is not $0$ and that $\varphi$ is in DNF. Furthermore, suppose $\bar{x} = (x_1,\ldots,x_{\length(\bar{x})})$. Then $\alpha = \sa{\bar{x}} \varphi_1(\bar{x}) \vee \cdots \vee \varphi_n(\bar{x})$. Since $\alpha$ is not null, then some $\varphi_i$ must be satisfiable. This is, the function defined by $\sa{\bar{x}}\varphi(\bar{x})$ is not null. We will prove by induction on $\length(\bar{x})$ that the function defined by $\sa{\bar{x}}\varphi(\bar{x})$ is either greater or equal to $n$, or in $\Omega(n^2)$. We address the case $\length(\bar{x})= 1$, then $\alpha = \sa{x}\bigwedge\psi(x)$. If any $\psi(x) = (x = x)$ or $\neg(x < x)$, then we can eliminate it and we obtain the same function. If any $\psi = (x < x)$ or $\neg(x=x)$, then the function becomes null. If $\psi(x) = R(x,\ldots,x)$ for some $R\in\R$ the function becomes null for the structures we are considering. If $\psi(x) = \neg R(x,\ldots,x)$, we can eliminate it and for the structures we are considering we obtain the same function. The only possible $\alpha$ left is $\alpha = \sa{x}\top$ which is equal to the function $n$. This covers all possible cases for $\length(\bar{x}) = 1$. Now suppose that it holds for $\length(\bar{x}) = k$ and suppose $\alpha = \sa{\bar{x}}\bigwedge\psi(\bar{x})$ for $\length(\bar{x}) = k+1$. If any $\psi(\bar{x}) = (x_i = x_j)$ where $i \neq j$, then $\alpha$ describes the same function as $\alpha$ where $x_j$ has been replaced by $x_i$. In this formula the tuple of first-order variables has $k$ elements so the function it describes if one of the mentioned in the hypothesis. If $i = j$, then we can eliminate it and obtain the same function. If any $\psi(\bar{x}) = R(\bar{v})$ or $\neg R(\bar{v})$ where $\bar{v}$ is a sub-tuple of $\bar{x}$ then we can either eliminate it or the function becomes null, following the same argument as in the case $\length(\bar{x}) = 1$. If any $\psi(\bar{x}) = \neg(x_i = x_j)$ or $(x_i < x_j)$ where $i = j$, then the function becomes null. If any $\psi(\bar{x}) = \neg(x_i < x_j)$ where $i = j$, we can eliminate it. The remaining formulas in $\bigwedge\psi(\bar{x})$ are either $\neg(x_i = x_j)$, $(x_i<x_j)$ or $\neg(x_i<x_j)$. If the formula violates transitivity in $<$ (for example, $x < y \wedge y < z \wedge z < x$), then the function $\alpha$ describes is null. Therefore, there is some order over $\bar{x}$ that satisfies $\bigwedge\psi(\bar{x})$. Consider the formula that describes this order (like $x_1 < x_3 \wedge x_3 < x_4 \wedge x_4 < x_2$). The function $\alpha$ describes is greater or equal to the one this formula describes, which is exactly $\binom{n}{\length(\bar{x})}$ which is in $\Omega(n^{\length(\bar{x})}) \subseteq \Omega(n^2)$ if $\length(\bar{x}) > 1$. This concludes the proof of the claim.
We suppose that $\alpha$ is in SNF, this is, $\alpha = c + \sum_{i = 1}^n\alpha_i$. Let $\mathbb{1}$ be a $\R$-structure with one element. Note that if $c > 0$, then $\alpha(\enc(\mathbb{1})) \geq c > 0$ which is not possible since $\alpha(\enc(\mathbb{1})) = 0$. Since $\alpha$ is not null, consider some $\alpha_i$ that describes a non-null function. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$, where $\varphi$ is quantifier-free. Note that if $\length(\bar{X}) > 0$, then the function $\alpha$ describes is in $\Omega(2^n)$, as it was proven by the authors in \cite{SalujaST95}. We have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$, as we proved in the claim, describes either some function greater or equal to $n$, or in $\Omega(n^2)$, which leads to a contradiction. Lastly, note that the formula $\sa{x}\exists y(x < y)$ is in $\E{0}$ and describes the function $n-1$, which concludes the proof.

\vspace{1em}
We show here that $\E{1}\not\subseteq\QE{0}$. By contradiction, suppose that the constant 1 is expressible in $\E{1}$. Let $\alpha = \sa{\bar{X}}\sa{\bar{x}}\exists\bar{y}\,\varphi(\bar{X},\bar{x},\bar{y})$ describe this function for some quantifier-free formula $\varphi$.

### 2016-12-31 LICS/syntactic-horn.tex
31 dic merge

\begin{proposition}\label{prop:hsat-not-sigma2}	
$\chsat \not\in \eqso(\loge{2})$.
\end{proposition}

We conclude this section by showing that $\eqso(\ehorn)$ is the class we were looking for, as not only every function in $\eqso(\ehorn)$ has an easy decision counterpart, but also $\eqso(\ehorn)$ admits a natural complete problem under parsimonious reductions. More precisely, define $\dhsat$ as the decision problem:\begin{multline*}
\{\Phi \mid \Phi \text{ is a disjunction of}\\  \text{Horn formulas and $\Phi$ is satisfiable}\},
\end{multline*}
and define $\shdhsat$ as the problem of counting the satisfying assignments of a formula $\Phi$ that is a disjunction of Horn formulas. Then we have that:

>>>>

\begin{proposition}\label{prop:hsat-not-sigma2}	
$\chsat \not\in \eqso(\loge{2})$.
\end{proposition}
We conclude this section by showing that $\eqso(\ehorn)$ is the class we were looking for, as not only every function in $\eqso(\ehorn)$ has an easy decision counterpart, but also $\eqso(\ehorn)$ admits a natural complete problem under parsimonious reductions. More precisely, define $\dhsat$ as the decision problem:
\begin{multline*}
\{\Phi \mid \Phi \text{ is a disjunction of}\\  \text{Horn formulas and $\Phi$ is satisfiable}\},
\end{multline*}
and define $\shdhsat$ as the problem of counting the satisfying assignments of a formula $\Phi$ that is a disjunction of Horn formulas. Then we have that:

### 2016-12-31 QSO/cap-comp.tex
31 dic merge

	We represent configurations with a tuple of fixed size $k$. The formula $\varphi(\bar{x},\bar{y})$ describes a procedure that given a configuration generates a possible next configuration. The formula $\varphi_I(\bar{x})$ models that $\bar{x}$ is the initial configuration of $M(w)$. The formula $\varphi_F(\bar{x})$ models that $\bar{x}$ is an accepting (final) configuration of $M(w)$. The formula we construct is:
	$$
	\alpha = \sa{\bar{x}}\sa{\bar{y}}([\pth \varphi(\bar{u},\bar{v})](\bar{x},\bar{y})\cdot \varphi_I(\bar{x})\cdot\varphi_F(\bar{y})).
	$$

	To illustrate our idea, we will show a greatly simplified example. Consider a machine $M$ that works in exactly $\log_2(n)$ space and only allows 0 or 1 in the working tape. Consider an input $\A$ of size 16 (that is, $A = \{0,\ldots,9,A,\ldots,F\}$). Let some configuration $s$ have 0011 in the working tape, the head in the input tape is in position 26, and the head in the input tape is in position 2 (we consider 0-indexed positions). Also, $Q = \{q_1,\ldots,q_5\}$ and the current state is $q_3$.

@@

	The problem that arises from this representation, is that to describe a transition in $M$ we need to read an arbitrary character in the working tape, and therefore we have to obtain the $a_3$-th bit in $a_4$. Furthermore, to get the following configuration, we need compute $a_4$ with the $a_3$-th bit flipped. This is generally not possible to describe with an $FO$ formula. Consider the following procedure that receives $x = a_4$ and $i = a_3$.

	\begin{algorithm} \label{switch1to0}
	\caption{If the $i$-th bit in $x$ is 1 replace it by 0 and return the result}
	\begin{algorithmic}
		\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
		\While{$j > 0$}
		\State $v \gets 0$
		\While{$u > 1$}
		\State $u \gets u-2,\; v \gets v+1$
		\EndWhile
		\State $u\gets v,\; j \gets j-1$
		\EndWhile
		\While{$u > 1$}
		\State $u \gets u-2$
		\EndWhile
		\State $\textbf{assert } u = 1$ \Comment{If $u \neq 1$ simply stop}	
		\State $y \gets 1$ \Comment{Compute $2^i$ and store it in $y$}
		\While{$i > 0$}
		\State $z \gets 0$
		\While{$y > 0$}
		\State $z \gets z+2,\; y \gets y-1$
		\EndWhile
		\State $i \gets i-1,\; y \gets z$
		\EndWhile
		\While{$y > 0$} \Comment{Substract $y$ from $x$}
		\State $x \gets x-1,\; y \gets y-1$
		\EndWhile
		\State \Return $x$.
	\end{algorithmic}
	\end{algorithm}	
	Each of the instructions can be expressed with $\fo$, so our strategy is to use the $\pth$ operator to simulate the algorithm and then we can describe a transition using the processed value of $a_4$. Note that we need to describe three more procedures to simulate the transitions for $0\to 0$, $0 \to 1$ and $1\to 1$.

@@

	Note that we also need to specify that the program we are following is Algorithm \ref{switch1to0} so we store $1,0$ in $b_1,b_2$ all along the procedure. We now describe the three other algorithms that compute the switches from $0\to 0$, $0\to 1$ and $1\to 1$.
	\begin{algorithm} \label{switch0to0}
		\caption{If the $i$-th bit in $x$ is 0 return $x$}
		\begin{algorithmic}
			\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
			\While{$j > 0$}
			\State $v \gets 0$
			\While{$u > 1$}
			\State $u \gets u-2,\; v \gets v+1$
			\EndWhile
			\State $u\gets v,\; j \gets j-1$
			\EndWhile
			\While{$u > 1$}
			\State $u \gets u-2$
			\EndWhile
			\State $\textbf{assert } u = 0$ \Comment{If $u \neq 0$ simply stop}	
			\State \Return $x$.
		\end{algorithmic}
	\end{algorithm}
	\begin{algorithm} \label{switch0to1}
		\caption{If the $i$-th bit in $x$ is 0 replace it by 1 and return the result}
		\begin{algorithmic}
			\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
			\While{$j > 0$}
			\State $v \gets 0$
			\While{$u > 1$}
			\State $u \gets u-2,\; v \gets v+1$
			\EndWhile
			\State $u\gets v,\; j \gets j-1$
			\EndWhile
			\While{$u > 1$}
			\State $u \gets u-2$
			\EndWhile
			\State $\textbf{assert } u = 0$ \Comment{If $u \neq 0$ simply stop}	
			\State $y \gets 1$ \Comment{Compute $2^i$ and store it in $y$}
			\While{$i > 0$}
			\State $z \gets 0$
			\While{$y > 0$}
			\State $z \gets z+2,\; y \gets y-1$
			\EndWhile
			\State $i \gets i-1,\; y \gets z$
			\EndWhile
			\While{$y > 0$} \Comment{Add $y$ to $x$}
			\State $x \gets x+1,\; y \gets y-1$
			\EndWhile
			\State \Return $x$.
		\end{algorithmic}
	\end{algorithm}	
	\begin{algorithm} \label{switch1to1}
		\caption{If the $i$-th bit in $x$ is 1 return $x$}
		\begin{algorithmic}
			\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
			\While{$j > 0$}
			\State $v \gets 0$
			\While{$u > 1$}
			\State $u \gets u-2,\; v \gets v+1$
			\EndWhile
			\State $u\gets v,\; j \gets j-1$
			\EndWhile
			\While{$u > 1$}
			\State $u \gets u-2$
			\EndWhile
			\State $\textbf{assert } u = 1$ \Comment{If $u \neq 0$ simply stop}	
			\State \Return $x$.
		\end{algorithmic}
	\end{algorithm}
	For the other three cases, where $\delta = (q_i,a,0,q_j,op_1,0,op_2)$, $\delta = (q_i,a,0,q_j,op_1,1,op_2)$ and $\delta = (q_i,a,1,q_j,op_1,1,op_2)$, $\varphi_{\delta}(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v})$ is defined analogously. Then, $\varphi(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v})$ is defined as:
	$$
	\varphi(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v}) = \bigvee_{\delta \in \Delta} \varphi_{\delta}(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v}).
	$$
	Finally we define $\varphi_I$ and $\varphi_F$:
	\begin{align*}
	\varphi_I(\bar{x},\bar{s},\bar{w},\bar{u}) &= \varphi_0(x_1) \wedge \varphi_0(x_2) \wedge \varphi_0(x_3) \wedge \bigwedge_{i = 4}^{3+c}\varphi_0(x_i) \wedge \varphi_1(x_{4+c}) \wedge \bigwedge_{i = 5+c}^{3+c+\ell} \varphi_0(x_i)\wedge \varphi_{0,0}(s_1,s_2)\ \wedge \varphi^b_0(\bar{w}) \wedge \bigwedge_{i = 1}^{5c+2}\varphi_0(u_i). \\
	\varphi_F(\bar{x},\bar{s},\bar{w},\bar{u}) &= \varphi^q_{\ell}(x_{4+c},\ldots,x_{3+c+\ell}) \wedge \varphi_{0,0}(s_1,s_2)\ \wedge \varphi^b_0(\bar{w}) \wedge \bigwedge_{i = 1}^{5c+2}\varphi_0(u_i),
	\end{align*}
	and then $\sem{\alpha}(\A) = \sem{\sa{\bar{x}}\sa{\bar{y}}([\pth \varphi(\bar{u},\bar{v})](\bar{x},\bar{y})\cdot \varphi_I(\bar{x})\cdot\varphi_F(\bar{y}))}(\A) = \acc_M(\A)$.

@@

%\begin{theorem}
%$\eqso(\integ,\fo)$ captures $\gp$ over the class of ordered structures.
%\end{theorem}

>>>>

	We represent configurations with a tuple of fixed size. The formula $\varphi(\bar{x},\bar{y})$ describes a procedure that given a configuration generates a possible next configuration. The formula $\varphi_I(\bar{x})$ describes that $\bar{x}$ is the initial configuration of $M(w)$. The formula $\varphi_F(\bar{x})$ describes that $\bar{x}$ is an accepting (final) configuration of $M(w)$. The formula we construct is:
	$$
	\alpha = \sa{\bar{x}}\sa{\bar{y}}([\pth \varphi(\bar{u},\bar{v})](\bar{x},\bar{y})\cdot \varphi_I(\bar{x})\cdot\varphi_F(\bar{y})).
	$$

	To illustrate our idea, we will show a simplified example. Consider a machine $M$ that works in exactly $\log_2(n)$ space and only allows 0 or 1 in the working tape. Consider an input $\A$ of size 16 (that is, $A = \{0,\ldots,9,A,\ldots,F\}$). Let some configuration $s$ have 0011 in the working tape, the head in the input tape is in position 26, and the head in the input tape is in position 2 (we consider 0-indexed positions). Also, $Q = \{q_1,\ldots,q_5\}$ and the current state is $q_3$.

@@

	The problem that arises from this representation, is that to describe a transition in $M$ we need to read an arbitrary character in the working tape, and therefore we have to obtain the $a_3$-th bit in $a_4$. Furthermore, to represent the following configuration, we need compute $a_4$ with the $a_3$-th bit flipped. This is generally not possible to describe with an $\fo$ formula. Consider the following procedure that receives $x = a_4$ and $i = a_3$.

	\begin{algorithm}
	\caption{If the $i$-th bit in $x$ is 1 replace it by 0 and return the result}
	\label{switch1to0}
	\begin{algorithmic}
		\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
		\While{$j > 0$}
		\State $v \gets 0$
		\While{$u > 1$}
		\State $u \gets u-2,\; v \gets v+1$
		\EndWhile
		\State $u\gets v,\; j \gets j-1$
		\EndWhile
		\While{$u > 1$}
		\State $u \gets u-2$
		\EndWhile
		\State $\textbf{assert } u = 1$ \Comment{If $u \neq 1$ simply stop}	
		\State $y \gets 1$ \Comment{Compute $2^i$ and store it in $y$}
		\While{$i > 0$}
		\State $z \gets 0$
		\While{$y > 0$}
		\State $z \gets z+2,\; y \gets y-1$
		\EndWhile
		\State $i \gets i-1,\; y \gets z$
		\EndWhile
		\While{$y > 0$} \Comment{Substract $y$ from $x$}
		\State $x \gets x-1,\; y \gets y-1$
		\EndWhile
		\State \Return $x$.
	\end{algorithmic}
	\end{algorithm}	
	Each of the instructions can be expressed with $\fo$, so our strategy is to use the $\pth$ operator to simulate the algorithm and then we can describe a transition using the processed value of $a_4$. Note that we need to describe three more procedures to simulate the transitions for $0\to 0$, $0 \to 1$ and $1\to 1$.

@@

	\begin{algorithm}
		\caption{If the $i$-th bit in $x$ is 0 return $x$} \label{switch0to0}
		\begin{algorithmic}
			\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
			\While{$j > 0$}
			\State $v \gets 0$
			\While{$u > 1$}
			\State $u \gets u-2,\; v \gets v+1$
			\EndWhile
			\State $u\gets v,\; j \gets j-1$
			\EndWhile
			\While{$u > 1$}
			\State $u \gets u-2$
			\EndWhile
			\State $\textbf{assert } u = 0$ \Comment{If $u \neq 0$ simply stop}	
			\State \Return $x$.
		\end{algorithmic}
	\end{algorithm}

	\begin{algorithm}
		\caption{If the $i$-th bit in $x$ is 0 replace it by 1 and return the result}
		\label{switch0to1}
		\begin{algorithmic}
			\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
			\While{$j > 0$}
			\State $v \gets 0$
			\While{$u > 1$}
			\State $u \gets u-2,\; v \gets v+1$
			\EndWhile
			\State $u\gets v,\; j \gets j-1$
			\EndWhile
			\While{$u > 1$}
			\State $u \gets u-2$
			\EndWhile
			\State $\textbf{assert } u = 0$ \Comment{If $u \neq 0$ simply stop}	
			\State $y \gets 1$ \Comment{Compute $2^i$ and store it in $y$}
			\While{$i > 0$}
			\State $z \gets 0$
			\While{$y > 0$}
			\State $z \gets z+2,\; y \gets y-1$
			\EndWhile
			\State $i \gets i-1,\; y \gets z$
			\EndWhile
			\While{$y > 0$} \Comment{Add $y$ to $x$}
			\State $x \gets x+1,\; y \gets y-1$
			\EndWhile
			\State \Return $x$.
		\end{algorithmic}
	\end{algorithm}

	\begin{algorithm}
		\caption{If the $i$-th bit in $x$ is 1 return $x$}
		\label{switch1to1}
		\begin{algorithmic}
			\State $u \gets x,\; j \gets i$ \Comment{Get the $i$-th bit on $x$ and store it in $u$}
			\While{$j > 0$}
			\State $v \gets 0$
			\While{$u > 1$}
			\State $u \gets u-2,\; v \gets v+1$
			\EndWhile
			\State $u\gets v,\; j \gets j-1$
			\EndWhile
			\While{$u > 1$}
			\State $u \gets u-2$
			\EndWhile
			\State $\textbf{assert } u = 1$ \Comment{If $u \neq 0$ simply stop}	
			\State \Return $x$.
		\end{algorithmic}
	\end{algorithm}
	Note that we also need to specify that the program we are following is Algorithm \ref{switch1to0} so we store $1,0$ in $b_1,b_2$ all along the procedure. We describe the three other algorithms that compute the switches from $0\to 0$, $0\to 1$ and $1\to 1$ (Algorithms \ref{switch0to0}, \ref{switch0to1} and \ref{switch1to1}.)
	For the other three cases, where $\delta = (q_i,a,0,q_j,op_1,0,op_2)$, $\delta = (q_i,a,0,q_j,op_1,1,op_2)$ and $\delta = (q_i,a,1,q_j,op_1,1,op_2)$, $\varphi_{\delta}(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v})$ is defined analogously. Then, $\varphi(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v})$ is defined as:
	$$
	\varphi(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v}) = \bigvee_{\delta \in \Delta} \varphi_{\delta}(\bar{x},\bar{s},\bar{w},\bar{u},\bar{y},\bar{t},\bar{z},\bar{v}).
	$$
	Finally we define $\varphi_I$ and $\varphi_F$:
	\begin{align*}
	\varphi_I(\bar{x},\bar{s},\bar{w},\bar{u}) &= \varphi_0(x_1) \wedge \varphi_0(x_2) \wedge \varphi_0(x_3) \wedge \bigwedge_{i = 4}^{3+c}\varphi_0(x_i) \wedge \varphi_1(x_{4+c}) \wedge \bigwedge_{i = 5+c}^{3+c+\ell} \varphi_0(x_i)\wedge \varphi_{0,0}(s_1,s_2)\ \wedge \varphi^b_0(\bar{w}) \wedge \bigwedge_{i = 1}^{5c+2}\varphi_0(u_i). \\
	\varphi_F(\bar{x},\bar{s},\bar{w},\bar{u}) &= \varphi^q_{\ell}(x_{4+c},\ldots,x_{3+c+\ell}) \wedge \varphi_{0,0}(s_1,s_2)\ \wedge \varphi^b_0(\bar{w}) \wedge \bigwedge_{i = 1}^{5c+2}\varphi_0(u_i),
	\end{align*}
	and then $\sem{\alpha}(\A) = \sem{\sa{\bar{x}}\sa{\bar{y}}([\pth \varphi(\bar{u},\bar{v})](\bar{x},\bar{y})\cdot \varphi_I(\bar{x})\cdot\varphi_F(\bar{y}))}(\A) = \acc_M(\A)$.

@@

\begin{theorem}
$\eqso(\integ,\fo)$ captures $\gp$ over the class of ordered structures.
\end{theorem}
